Double White Dwarf Mergers with CASTRO\\ I. Methodology and Code Verification

### 2015-02-18 papers/wdmerger_I/paper.tex
Add plot for the boundary condition comparison to the paper.

Analytic solutions to the Poisson equation customarily assume that the
potential vanishes at large distances from the region of non-zero
density. However, on a finite computational domain it is usually not
possible to have the edges of the domain be far enough away that the
potential can be taken to be zero there. Solving the Poisson equation
therefore requires knowledge of the values of the potential on the
edges of the computational domain. In principle, they can be computed
by doing a direct sum over the mass distribution inside the domain,
where the mass in each zone is treated as a point source:
\begin{equation}
  \Phi_{{lmn}} = -\sum_{{i, j, k}} \frac{G \rho_{{ijk}}}{|\mathbf{x}_{{lmn}} - \mathbf{x}_{{ijk}}|}\, \Delta V_{{ijk}}.\label{direct_sum}
\end{equation}
Here $(i, j, k)$ are the indices of cells inside the domain, and $(l,m, n)$ 
are the indices of boundary locations. $\Delta V$ is the volume of the
zone. We have implemented this as an option\footnote{It is controlled
  with the \texttt{gravity.direct\_sum\_bcs} input parameter.} in
CASTRO. If there are $N$ zones per spatial dimension, then there are
$6 N^2$ boundary zones, and each boundary zone requires a sum over
$N^3$ zones, so the direct computation of the boundary conditions
scales as $\mathcal{O}(N^5)$.  This method is expensive enough that it is not used
for hydrodynamics simulations (though it is useful for comparison to
approximate solutions).

@@

% Insert multipole vs. exact calculation here

@@

Since the results of a merger simulation depend strongly on gravity,
it is important to check whether proper numerical convergence is
achieved for the Poisson solver. To do so, we created a simple test
that inserts a sphere of radius $R$ and uniform mass density $\rho$
onto our grid, and used CASTRO to calculate the gravitational
potential $\Phi$ of this setup. We ensure that $R$ is an integer
multiple of the grid spacing, and the center of the sphere is at the
origin. The problem domain for our simulations is $[-1.6, 1.6]^3$, and
we take $R = 0.5$ and $\rho = 10^3$. The zones with $r > R$ are filled
with an ambient material of very low density. We run this problem at
multiple resolutions corresponding to jumps by a factor of two. For
comparison, at each grid point we evaluate the analytical potential of
a uniform sphere, which can be easily determined using Gauss' law:
\begin{equation}
  \Phi_{\text{sphere}}(r) = -\frac{GM}{r} \times \begin{cases} (3R^2 - r^2)/(2 r^2) & r \leq R \\ 1 & r > R \end{cases},\label{eq:sphere-analytical}
\end{equation}
where $M = 4\pi R^3 / 3$ is the mass of the sphere. Defining the $L^2$
norm of a field $f$ as
\begin{equation}
  \| f \|_2 = \left(\sum_{i,j,k} dx\, dy\, dz\, f_{ijk}^2\right)^{1/2},
\end{equation}
we measure the numerical error by calculating the $L^2$ norm of the error and normalizing it by the $L^2$ norm of the analytical solution:
\begin{equation}
  \text{Error} = \frac{\|\Phi - \Phi_{\text{sphere}}\|_2}{\|\Phi_{\text{sphere}}\|_2}.
\end{equation}

@@

\begin{figure}
  \centering
  \includegraphics[scale=1.0]{reference}
  \caption{ \label{Fig:sod} Solution to Sod's problem with the original
    reference state and the new reference state, as compared to the
    exact solution.  We note that the new reference state shows a
    slightly sharper shock and contact, but also has a dip at the tail
    of the rarefaction.}
\end{figure}

@@

%\begin{figure}
%  \centering
%  \includegraphics[width=3.1in]{nested2}\hspace{1em}
%  \includegraphics[width=3.1in]{nested3}
%  \caption{\label{fig:2levgrav} Schematic showing the solve for $\phi$ for a 2-level grid.}
%\end{figure}

>>>>

Analytic solutions to the Poisson equation customarily assume that the
potential vanishes at large distances from the region of non-zero
density. However, on a finite computational domain it is usually not
possible to have the edges of the domain be far enough away that the
potential can be taken to be zero there. Solving the Poisson equation
therefore requires knowledge of the values of the potential on the
edges of the computational domain. In principle, they can be computed
by doing a direct sum over the mass distribution inside the domain,
where the mass in each zone is treated as a point source:
\begin{equation}
  \Phi_{{lmn}} = -\sum_{{i, j, k}} \frac{G \rho_{{ijk}}}{|\mathbf{x}_{{lmn}} - \mathbf{x}_{{ijk}}|}\, \Delta V_{{ijk}}.\label{eq:direct_sum}
\end{equation}
Here $(i, j, k)$ are the indices of cells inside the domain, and $(l,m, n)$ 
are the indices of boundary locations. $\Delta V$ is the volume of the
zone. We have implemented this as an option\footnote{It is controlled
  with the \texttt{gravity.direct\_sum\_bcs} input parameter.} in
CASTRO. If there are $N$ zones per spatial dimension, then there are
$6 N^2$ boundary zones, and each boundary zone requires a sum over
$N^3$ zones, so the direct computation of the boundary conditions
scales as $\mathcal{O}(N^5)$.  This method is expensive enough that it is not used
for hydrodynamics simulations (though it is useful for comparison to
approximate solutions).

@@

As a demonstration of the method's accuracy, we consider the case of two 
white dwarfs of mass ratio 2/3, loaded onto the grid by the method described 
in Section \ref{sec:initial_models}. We terminated the simulation just after
initialization, so that we perform only an initial Poisson solve for this 
density distribution. We did this for values of $l_{\text{max}}$ ranging
from 0 to 20, and we also did this using the numerically exact solution 
provided by Equation \ref{eq:direct_sum}.  Defining the $L^2$
norm of a field $f$ as
\begin{equation}
  \| f \|_2 = \left(\sum_{i,j,k} dx\, dy\, dz\, f_{ijk}^2\right)^{1/2},
\end{equation}
we computed the $L^2$ error of $\Phi$ on the entire domain for multipole 
boundary conditions, which we will call $\Phi_l$, relative to $\Phi$ 
obtained using the exact boundary conditions:
\begin{equation}
  \text{Error}_l = \frac{\|\Phi_l - \Phi_{\text{exact}}\|_2}{\|\Phi_{\text{exact}}\|_2}.
\end{equation}
The result is shown in Figure \ref{fig:bc_comparison}. At $l_{\text{max}} = 6$,
the error is already well below $10^{-4}$ and we adopt this as our default 
choice for all simulations with Poisson gravity. Our choice is also motivated by the 
fact that at this value, the time for computing the boundary conditions on the
coarse grid is comparable to the time for the multigrid solve itself; at higher values 
the boundary condition calculation becomes much more expensive than the multigrid solve
and so the overall returns are severely diminished. We note that in our implementation 
there are no gains at all for going beyond $l_{\text{max}} \gtrsim 16$, as we 
become limited by error due to the machine precision.
\begin{figure}[h]
  \centering
  \includegraphics[scale=0.45]{plots/bc_comparison}
  \caption{Error of $\Phi$ on the computational domain for a binary white dwarf simulation 
    whose boundary conditions were computed using various values of the maximum multipole order,
    relative to the exact solution determined by a brute force sum on the boundaries.
    Circles represent the error at integer values, and they have been connected by a smooth 
    line to guide the eye.\label{fig:bc_comparison}}
\end{figure}

@@

Since the results of a merger simulation depend strongly on gravity,
it is important to check whether proper numerical convergence is
achieved for the Poisson solver. To do so, we created a simple test
that inserts a sphere of radius $R$ and uniform mass density $\rho$
onto our grid, and used CASTRO to calculate the gravitational
potential $\Phi$ of this setup. We ensure that $R$ is an integer
multiple of the grid spacing, and the center of the sphere is at the
origin. The problem domain for our simulations is $[-1.6, 1.6]^3$, and
we take $R = 0.5$ and $\rho = 10^3$. The zones with $r > R$ are filled
with an ambient material of very low density. We run this problem at
multiple resolutions corresponding to jumps by a factor of two. For
comparison, at each grid point we evaluate the analytical potential of
a uniform sphere, which can be easily determined using Gauss' law:
\begin{equation}
  \Phi_{\text{sphere}}(r) = -\frac{GM}{r} \times \begin{cases} (3R^2 - r^2)/(2 r^2) & r \leq R \\ 1 & r > R \end{cases},\label{eq:sphere-analytical}
\end{equation}
where $M = 4\pi R^3 / 3$ is the mass of the sphere. We measure the 
numerical error by calculating the $L^2$ norm of the error and 
normalizing it by the $L^2$ norm of the analytical solution:
\begin{equation}
  \text{Error} = \frac{\|\Phi - \Phi_{\text{sphere}}\|_2}{\|\Phi_{\text{sphere}}\|_2}.
\end{equation}

@@

\begin{figure}
  \centering
  \includegraphics[scale=1.0]{plots/reference}
  \caption{ \label{Fig:sod} Solution to Sod's problem with the original
    reference state and the new reference state, as compared to the
    exact solution.  We note that the new reference state shows a
    slightly sharper shock and contact, but also has a dip at the tail
    of the rarefaction.}
\end{figure}

@@

%\begin{figure}
%  \centering
%  \includegraphics[width=3.1in]{plots/nested2}\hspace{1em}
%  \includegraphics[width=3.1in]{plots/nested3}
%  \caption{\label{fig:2levgrav} Schematic showing the solve for $\phi$ for a 2-level grid.}
%\end{figure}
